---
title: "CellLine_Br"
author: "Vagmita Pabuwal"
date: "5/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This R script is to analyze general information about cell line data without any experimental effect. This can be used to understand differential count and clustering of cell lines and genes if any experimental condition would 
have been applied on the cell lines
This Analysis primarily was done to understand the effect of different therapeutic agent on different Breast cancer cell lines.There are 30 cell lines with 6 different condition, no replicates. 
I have taken only the TPM values of expression without any therapeutic agent given.
Aim was to see based on TPM data how differnt are the cell lines expression. With the aim top 100 genes were plotted as a heatmap and differential count was calculated for one pair. We can calculate pairwise in similar way for all other cell lines.
Just to show few statistical analysis performance, I performed PCA and cluster analysis to see if the cell lines could be differentiates, One cell line definitely stood out which was HCC1806 (squamous cell breast carcinoma, acantholytic variant)
Overall we can further analyze data based on the data to see effect of drug on the cell lines and comparing the variant data to check which genes are most affected. 

#Setting the directory and loading required libraries
libraries can also be loaded or installed as per requirement while doing our analysis

```{r message=FALSE, warning=FALSE}

setwd("/Users/vagmi/Documents/")
library(edgeR)
library(RColorBrewer)
library(scatterplot3d)
library(dplyr)
library(DESeq2)
library(ggplot2)
library(plyr)
library(gplots)
library(pheatmap)
library(stats)
library(ggplot2)
library(ggfortify)
library(factoextra)

```
Reading the data and cleaning a bit
There were 2 duplicate genes which were removed
Wanted to see if there were lot of mitochondrial gene, but they were not as may so kept all in the data.

```{r}
Data_Br <- read.csv("/Users/vagmi/Documents/EMTAB_BreastCL.csv", header=TRUE, stringsAsFactors = TRUE,  row.names = 1)
Data_Br_unique <- Data_Br[!duplicated(Data_Br$Gene.Name), ]

rownames(Data_Br_unique) <- Data_Br_unique[,1]
Data_Br_unique <- Data_Br_unique[,-1]
count_Br_unique <- Data_Br_unique[rowSums(Data_Br_unique >20) >=1,]
mito_gene <- count_Br_unique[grep("^MT-", rownames(count_Br_unique)),]
Exp_design <- read.csv("Exp_design.csv", header=T)
Exp_design <- Exp_design[,-1]
colData <- read.csv("Exp_design.csv", header=T, stringsAsFactors = TRUE)
colData <- colData[,-1]
designFormula <- "~group"

```
plot Heatmap and PCA based on variance #Transpose the matrix for PCA ##Transforming to log2 scale #COmputing PCA plot PCA using ggplot2

```{r}
V <- apply(round(Data_Br_unique), 1, var)
selectedGenes <- names(V[order(V, decreasing = T)][1:80])

pheatmap(as.matrix(Data_Br_unique[selectedGenes,]))

Matrix <- t(Data_Br_unique[selectedGenes,])

Matrix <- log2(Matrix+1)

pcaResults <- prcomp(Matrix)

autoplot(pcaResults, data = colData)

```


Another way of plotting PCA


```{r}

colNames1 = colnames(Data_Br_unique)
colorInterpolation = colorRampPalette(brewer.pal(9,'Set1'))
col1 = colorInterpolation(length(sort(unique(colNames1))))
names(col1) = sort(unique(colNames1))
cols1 = as.character(col1[colNames1])



plot(pcaResults$x[,'PC1'],pcaResults$y[,'PC2'],main='PCA of cancer',col=cols1,pch=19,xlab='PC 1',ylab='PC 2')
legend('topright',legend=unique(colNames1),fill=col1)

```

DESeq prereq### ###Differential count but primarily to find genes based on pval  

```{r message=FALSE, warning=FALSE}



condition <- factor(c('invasive ductal carcinoma', 'breast adenocarcinoma', 'breast carcinoma', 'metaplastic breast carcinoma
','squamous cell breast carcinoma, acantholytic variant', 'breast ductal adenocarcinoma'))

dds <- DESeqDataSetFromMatrix(countData = round(Data_Br_unique), colData = colData, design =~ condition)
dds <- DESeq(dds) 
```


Compare 2 conditions and get gene based on pvalue for pairwise set

```{r}
DEresults_BACvsBC = results(dds, contrast = c("condition", 'breast adenocarcinoma','breast carcinoma'))

DEresults_BACvsBC <- DEresults_BACvsBC[order(DEresults_BACvsBC$pvalue),]
print (DEresults_BACvsBC)
write.csv(DEresults_BACvsBC, file="DEresults_BACvsBC.csv")

```

PCA & Plot heatmap for normalized DESEq result

```{r}
countsNormalized <- DESeq2::counts(dds, normalized=TRUE)
selectedGenes_dds <- names(sort(apply(countsNormalized, 1, var), decreasing=TRUE)[1:50])
normMatrix <- t(countsNormalized[selectedGenes_dds,])
normMatrix <- log2(normMatrix+1)
pcaResult_Norm<- prcomp(normMatrix)

autoplot(pcaResult_Norm, data = colData)

pheatmap(as.matrix(countsNormalized[selectedGenes_dds,]))


rld_Br <- rlog(dds)
DESeq2::plotPCA(rld_Br, ntop=100, intgroup = 'condition') + ylim(-25,25) + theme_bw()

```


CLuster using factoextra based on normalized count from DESeq


```{r}
countsNormalized_log <- log2(countsNormalized + 1)
scale_countsNormalized_log_t <- scale(t(countsNormalized_log))
km_countsNormalized_log_t <- kmeans(scale_countsNormalized_log_t, 6, nstart=25)
fviz_cluster(km_countsNormalized_log_t, scale_countsNormalized_log_t, ellipse = TRUE)
cluster_Br <- km_countsNormalized_log_t$cluster

```

cluster based on variance for top 100 genes

```{r}
scale_Matrix <- scale(Matrix)
km_Matrix <- kmeans(scale_Matrix, 6, nstart=25)
fviz_cluster(km_Matrix, scale_Matrix, ellipse = TRUE)
cluster_Br_Var <- km_Matrix$cluster

```

check the correlation between the cell lines

```{r}
correlationMatrix <- cor(countsNormalized)


corrplot::corrplot(correlationMatrix, order = 'hclust', 
                   addrect = 2, addCoef.col = 'white', 
                   number.cex = 0.7) 


pheatmap(correlationMatrix,cutree_cols = 2)

```

TSNE but data is not as huge

```{r}
library(Rtsne)
set.seed(46)
tsne.out <- Rtsne(Matrix,perplexity = 5)

plot(tsne.out$Y,col=as.factor(colData$condition),
     pch=19)

legend("bottomright",
       legend=unique(colData$condition),
       fill =palette("default"),
       border=NA,box.col=NA)
```
